Fractional Types

### 2019-11-09 fracGC/p.tex
a bit on paper

\usepackage{booktabs}  
\usepackage{subcaption} 
\usepackage{amssymb}
\usepackage{bbm}
\usepackage[greek,english]{babel}
\usepackage{ucs}
\usepackage[utf8x]{inputenc}
\usepackage{autofe}
\usepackage[references]{agda} 
\usepackage{newunicodechar}
\usepackage[nocenter]{qtree}
\usepackage{tikz}
\usetikzlibrary{cd}
\usetikzlibrary{quotes}

\newunicodechar{𝕌}{$\mathbb{U}$}
\newunicodechar{𝟙}{$1$}
\newunicodechar{ᵤ}{${}_u$}

\newcommand{\alt}{~|~}
\newcommand{\inlv}[1]{\ensuremath{\mathit{inl}(v)}}
\newcommand{\inrv}[1]{\ensuremath{\mathit{inr}(v)}}
\newcommand{\pointed}[2]{[#1 \bullet #2]}
\newcommand{\nboxtimes}[2]{\,\,~{^{#1}\boxtimes^{#2}}~\,\,}
\newcommand{\mm}{\texttt{\textminus}}
\newcommand{\pp}{\texttt{+}}
\newcommand{\inl}[1]{\textsf{inl}~#1}
\newcommand{\inr}[1]{\textsf{inr}~#1}
\newcommand{\idv}[3]{#2 \xrightarrow{#1} #3}
\newcommand{\cp}[3]{#1\stackrel{#2}{\bullet}#3}
\newcommand{\idt}[3]{#2 \equiv_{#1} #3}
\newcommand{\idrt}[3]{#3 \equiv_{#1} #2}
\newcommand{\refl}[1]{\textsf{refl}~#1}
\newcommand{\lid}{\textsf{lid}}
\newcommand{\rid}{\textsf{rid}}
\newcommand{\linv}{l!}
\newcommand{\rinv}{r!}
\newcommand{\invinv}{!!}
\newcommand{\assoc}{\circ}
\newcommand{\identlp}{\mathit{unite}_+\mathit{l}}
\newcommand{\identrp}{\mathit{uniti}_+\mathit{l}}
\newcommand{\identlsp}{\mathit{unite}_+\mathit{r}}
\newcommand{\identrsp}{\mathit{uniti}_+\mathit{r}}
\newcommand{\swapp}{\mathit{swap}_+}
\newcommand{\assoclp}{\mathit{assocl}_+}
\newcommand{\assocrp}{\mathit{assocr}_+}
\newcommand{\identlt}{\mathit{unite}_*\mathit{l}}
\newcommand{\identrt}{\mathit{uniti}_*\mathit{l}}
\newcommand{\identlst}{\mathit{unite}_*\mathit{r}}
\newcommand{\identrst}{\mathit{uniti}_*\mathit{r}}
\newcommand{\swapt}{\mathit{swap}_*}
\newcommand{\assoclt}{\mathit{assocl}_*}
\newcommand{\assocrt}{\mathit{assocr}_*}
\newcommand{\absorbr}{\mathit{absorbr}}
\newcommand{\absorbl}{\mathit{absorbl}}
\newcommand{\factorzr}{\mathit{factorzr}}
\newcommand{\factorzl}{\mathit{factorzl}}
\newcommand{\dist}{\mathit{dist}}
\newcommand{\factor}{\mathit{factor}}
\newcommand{\distl}{\mathit{distl}}
\newcommand{\factorl}{\mathit{factorl}}
\newcommand{\distz}{\mathit{absorbr}}
\newcommand{\iso}{\leftrightarrow}
\newcommand{\proves}{\vdash}
\newcommand{\idc}{\mathit{id}\!\!\leftrightarrow}
\newcommand{\Rule}[4]{
\makebox{{\rm #1}
$\displaystyle
\frac{\begin{array}{l}#2 \\\end{array}}
{\begin{array}{l}#3      \\\end{array}}$
 #4}}
\newcommand{\jdg}[3]{#2 \proves_{#1} #3}

@@

\url{https://www.nap.edu/read/25196/chapter/5#73}
\begin{quote}
  For error correction, one needs to replicate the state of a qubit
  onto a number of qubits. While the no cloning theorem prevents one
  from copying the state of a qubit directly onto another, one can
  create a redundant entangled qubit state of many qubits. The key is
  that the qubits to be entangled must start out in a known
  state. Qubits with a known state (for purposes of this discussion,
  it will be the state \verb.|0>.), called ``ancilla qubits,'' may be
  added to a computation for this purpose. Since the state of ancilla
  qubits are known, it is possible to create a simple circuit that
  makes the output state of all these ancilla qubits match the
  protected qubit: run each ancilla through a controlled-NOT gate,
  where the control is driven by the qubit that needs to be
  replicated. Assume that there is a qubit with state \verb.PSI. that
  we want to protect, where \verb.|PSI>. represents an arbitrary
  superposition state \verb.|PSI> = a0 |0> + a0 |1>. In the CNOT gate,
  the ancilla \verb.|0>. state will remain a \verb.|0>. state by the
  \verb.|0>. component of \verb.|PSI>., but it will be converted to
  \verb.|1>. by the \verb.|1>. component of \verb.|PSI>. The result of
  this operation is the newly entangled two-qubit state
  \verb.a0 |00> + a1 |11>., creating a system in which the ancilla
  qubit is now perfectly entangled with the first qubit. Adding more
  ancillas increases the distance of the repetition code.  \end{quote}

@@

\url{https://homepages.cwi.nl/~rdewolf/qcnotesv2.pdf}
\begin{quote}
  Using an ancilla qubit, it is possible to avoid doing any
  intermediate measurements in a quantum circuit. Show how this can be
  done.  Hint: instead of measuring the qubit, apply a CNOT that
  “copies” it to a new \verb.|0>.-qubit, which is then left alone
  until the end of the computation. Analyze what happens.
\end{quote}

@@

In-place matrix transpose: ease of programming, efficiency. Say we
want to transpose a matrix. Wikipedia example
(\url{https://en.wikipedia.org/wiki/In-place_matrix_transposition}):
\begin{verbatim}
11 12 13 14 
21 22 23 24 
\end{verbatim}
transposed to
\begin{verbatim}
11 21
12 22
13 23
14 24
\end{verbatim}
In Pi, the input and output matrices are:
\begin{verbatim}
M   = (11 , 21) , (12 , 22) , (13, 23) , (14 , 24)

trM = (11 , 12 , 13 , 14) , (21 , 22 , 23 , 23) 
\end{verbatim}
Say we are given a language like Pi that is sound and complete with
respect to permutations on finite types, we would write the
permutation like so.
\begin{verbatim}
WRITE PERMUTATION
\end{verbatim}
This code does not use additional space, i.e., it performs the matrix
transpose in constant space, i.e., performs an in-place matrix
transposition. It is well know that with additional space, one can
write more efficient matrix transpositions (explain and citations).

@@

Is the below a possible example?
\begin{verbatim}
Say we already have a permutation A <-> B
we can implement a permutation X <-> Z 
when there exists Y such that A = X * Y = Z * Y = B

X -> X * Y * 1/Y 
  -> A * 1/Y
  -> B * 1/Y
  -> Y * Z * 1/Y
  -> Z * Y * 1/Y
  -> Z
\end{verbatim}
This basically says, in the language of Lens, that
when $A$ is isomorphic to $X \times Y$ and
$B$ is isomorphic to $Z \times Y$, then
$X$ is isomorphic to $Y$.

@@

\begin{figure}[t]
\[\begin{array}{rrcll}
\idc :& \tau & \iso & \tau &: \idc \\
\\
\identlp :&  0 + \tau & \iso & \tau &: \identrp \\
\swapp :&  \tau_1 + \tau_2 & \iso & \tau_2 + \tau_1 &: \swapp \\
\assoclp :&  \tau_1 + (\tau_2 + \tau_3) & \iso & (\tau_1 + \tau_2) + \tau_3 &: \assocrp \\
\\
\identlt :&  1 * \tau & \iso & \tau &: \identrt \\
\swapt :&  \tau_1 * \tau_2 & \iso & \tau_2 * \tau_1 &: \swapt \\
\assoclt :&  \tau_1 * (\tau_2 * \tau_3) & \iso & (\tau_1 * \tau_2) * \tau_3 &: \assocrt \\
\\
\distz :&~ 0 * \tau & \iso & 0 ~ &: \factorzl \\
\dist :&~ (\tau_1 + \tau_2) * \tau_3 & \iso & (\tau_1 * \tau_3) + (\tau_2 * \tau_3)~ &: \factor
\end{array}\]
\begin{center}
\Rule{}
{\jdg{}{}{c_1 : \tau_1 \iso \tau_2} \quad \vdash c_2 : \tau_2 \iso \tau_3}
{\jdg{}{}{c_1 \odot c_2 : \tau_1 \iso \tau_3}}
{}
\\
\Rule{}
{\jdg{}{}{c_1 : \tau_1 \iso \tau_2} \quad \vdash c_2 : \tau_3 \iso \tau_4}
{\jdg{}{}{c_1 \oplus c_2 : \tau_1 + \tau_3 \iso \tau_2 + \tau_4}}
{}
\qquad
\Rule{}
{\jdg{}{}{c_1 : \tau_1 \iso \tau_2} \quad \vdash c_2 : \tau_3 \iso \tau_4}
{\jdg{}{}{c_1 \otimes c_2 : \tau_1 * \tau_3 \iso \tau_2 * \tau_4}}
{}
\end{center}
\caption{$\Pi$-terms and combinators.}
\label{pi-terms}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{PiFrac: Generalizing Scoped Ancilla, Runtime Check}

@@

When ancilla bits are created they are created with a specific
value. This value is encoded in the circuit. The value of fractional
type represents the ``garbage collector.''  The garbage collector is
specialized to collect a single value, which is the value used to
create the ancilla value. At this point, if the garbage collector
encounters a different value, it doesn't know what to do, and would
need to throw an exception. This situation in \verb|PiFracDyn.agda| is
similar to current solutions (e.g. Quipper) but actually more general
as it allows non-scoped ancilla allocation and de-allocation. Show
examples.

@@

%%%%%
\subsection{Ancilla}

@@

Exploit dependent types to reify proofs in the type system. Type
checking requires (partial) eval. Once we have a proof can we somehow
go back to \verb|PiFracDyn.agda| and remove the runtime check?

>>>>

\usepackage{booktabs}  
\usepackage{subcaption} 
\usepackage{amssymb}
\usepackage{stmaryrd}
\usepackage{bbm}
\usepackage[greek,english]{babel}
\usepackage{ucs}
\usepackage[utf8x]{inputenc}
\usepackage{autofe}
\usepackage[references]{agda} 
\usepackage{newunicodechar}
\usepackage[nocenter]{qtree}
\usepackage{tikz}
\usetikzlibrary{cd}
\usetikzlibrary{quotes}

\newunicodechar{𝕌}{$\mathbb{U}$}
\newunicodechar{𝔽}{$\mathbb{F}$}
\newunicodechar{𝕋}{$\mathbb{T}$}
\newunicodechar{𝟘}{$0$}
\newunicodechar{𝟙}{$1$}
\newunicodechar{𝟚}{$2$}
\newunicodechar{⋆}{${}_*$}
\newunicodechar{ᵤ}{${}_u$}
\newunicodechar{⊚}{$\fatsemi$}

\newcommand{\alt}{~|~}
\newcommand{\inlv}[1]{\ensuremath{\mathit{inl}(v)}}
\newcommand{\inrv}[1]{\ensuremath{\mathit{inr}(v)}}
\newcommand{\ket}[1]{|#1\rangle}
\newcommand{\pointed}[2]{[#1 \bullet #2]}
\newcommand{\nboxtimes}[2]{\,\,~{^{#1}\boxtimes^{#2}}~\,\,}
\newcommand{\mm}{\texttt{\textminus}}
\newcommand{\pp}{\texttt{+}}
\newcommand{\inl}[1]{\textsf{inl}~#1}
\newcommand{\inr}[1]{\textsf{inr}~#1}
\newcommand{\idv}[3]{#2 \xrightarrow{#1} #3}
\newcommand{\cp}[3]{#1\stackrel{#2}{\bullet}#3}
\newcommand{\idt}[3]{#2 \equiv_{#1} #3}
\newcommand{\idrt}[3]{#3 \equiv_{#1} #2}
\newcommand{\refl}[1]{\textsf{refl}~#1}
\newcommand{\lid}{\textsf{lid}}
\newcommand{\rid}{\textsf{rid}}
\newcommand{\linv}{l!}
\newcommand{\rinv}{r!}
\newcommand{\invinv}{!!}
\newcommand{\assoc}{\circ}
\newcommand{\identlp}{\mathit{unite}_+\mathit{l}}
\newcommand{\identrp}{\mathit{uniti}_+\mathit{l}}
\newcommand{\identlsp}{\mathit{unite}_+\mathit{r}}
\newcommand{\identrsp}{\mathit{uniti}_+\mathit{r}}
\newcommand{\swapp}{\mathit{swap}_+}
\newcommand{\assoclp}{\mathit{assocl}_+}
\newcommand{\assocrp}{\mathit{assocr}_+}
\newcommand{\identlt}{\mathit{unite}_*\mathit{l}}
\newcommand{\identrt}{\mathit{uniti}_*\mathit{l}}
\newcommand{\identlst}{\mathit{unite}_*\mathit{r}}
\newcommand{\identrst}{\mathit{uniti}_*\mathit{r}}
\newcommand{\swapt}{\mathit{swap}_*}
\newcommand{\assoclt}{\mathit{assocl}_*}
\newcommand{\assocrt}{\mathit{assocr}_*}
\newcommand{\absorbr}{\mathit{absorbr}}
\newcommand{\absorbl}{\mathit{absorbl}}
\newcommand{\factorzr}{\mathit{factorzr}}
\newcommand{\factorzl}{\mathit{factorzl}}
\newcommand{\dist}{\mathit{dist}}
\newcommand{\factor}{\mathit{factor}}
\newcommand{\distl}{\mathit{distl}}
\newcommand{\factorl}{\mathit{factorl}}
\newcommand{\distz}{\mathit{absorbr}}
\newcommand{\iso}{\leftrightarrow}
\newcommand{\proves}{\vdash}
\newcommand{\idc}{\mathit{id}\!\!\leftrightarrow}
\newcommand{\Rule}[4]{
\makebox{{\rm #1}
$\displaystyle
\frac{\begin{array}{l}#2 \\\end{array}}
{\begin{array}{l}#3      \\\end{array}}$
 #4}}
\newcommand{\jdg}[3]{#2 \proves_{#1} #3}

@@

\url{https://www.nap.edu/read/25196/chapter/5#73}
\begin{quote}
  For error correction, one needs to replicate the state of a qubit
  onto a number of qubits. While the no cloning theorem prevents one
  from copying the state of a qubit directly onto another, one can
  create a redundant entangled qubit state of many qubits. The key is
  that the qubits to be entangled must start out in a known
  state. Qubits with a known state (for purposes of this discussion,
  it will be the state $\ket{0}$), called ``ancilla qubits,'' may be
  added to a computation for this purpose. Since the state of ancilla
  qubits are known, it is possible to create a simple circuit that
  makes the output state of all these ancilla qubits match the
  protected qubit: run each ancilla through a controlled-NOT gate,
  where the control is driven by the qubit that needs to be
  replicated. Assume that there is a qubit with state $\Psi$ that we
  want to protect, where $\ket{\Psi}$ represents an arbitrary
  superposition state $\ket{\Psi} = a_0 \ket{0} + a_1 \ket{1}$. In the
  CNOT gate, the ancilla $\ket{0}$ state will remain a $\ket{0}$ state
  by the $\ket{0}$ component of $\ket{\Psi}$, but it will be converted
  to $\ket{1}$ by the $\ket{1}$ component of $\ket{\Psi}$. The result
  of this operation is the newly entangled two-qubit state
  $a_0 \ket{0} + a_1 \ket{1}$, creating a system in which the ancilla
  qubit is now perfectly entangled with the first qubit. Adding more
  ancillas increases the distance of the repetition code.  \end{quote}

@@

\url{https://homepages.cwi.nl/~rdewolf/qcnotesv2.pdf}
\begin{quote}
  Using an ancilla qubit, it is possible to avoid doing any
  intermediate measurements in a quantum circuit. Show how this can be
  done.  Hint: instead of measuring the qubit, apply a CNOT that
  “copies” it to a new $\ket{0}$-qubit, which is then left alone until
  the end of the computation. Analyze what happens.
\end{quote}

@@

In-place matrix transpose: ease of programming, efficiency. Say we
want to transpose a matrix. Wikipedia example
(\url{https://en.wikipedia.org/wiki/In-place_matrix_transposition}):
\begin{verbatim}
11 12 13 14 
21 22 23 24 
\end{verbatim}
transposed to
\begin{verbatim}
11 21
12 22
13 23
14 24
\end{verbatim}
In Pi, the input and output matrices are:
\begin{verbatim}
M = (11,21) , (12,22) , (13,23) , (14,24)

trM = (11,12,13,14) , (21,22,23,23) 
\end{verbatim}
Say we are given a language like Pi that is sound and complete with
respect to permutations on finite types, we would write the
permutation like so.
\begin{verbatim}
WRITE PERMUTATION
\end{verbatim}
This code does not use additional space, i.e., it performs the matrix
transpose in constant space, i.e., performs an in-place matrix
transposition. It is well know that with additional space, one can
write more efficient matrix transpositions (explain and citations).

@@

Is the below a possible example?

Say we already have a permutation $A \leftrightarrow B$
we can implement a permutation $X \leftrightarrow Z$ 
when there exists $Y$ such that $A = X * Y = Z * Y = B$:
\[\begin{array}{rcl}
X &\rightarrow& X * Y * 1/Y \\
  &\rightarrow& A * 1/Y \\
  &\rightarrow&  B * 1/Y \\
  &\rightarrow&  Y * Z * 1/Y \\
  &\rightarrow&  Z * Y * 1/Y \\
  &\rightarrow&  Z
\end{array}\]
This basically says, in the language of Lens, that
when $A$ is isomorphic to $X \times Y$ and
$B$ is isomorphic to $Z \times Y$, then
$X$ is isomorphic to $Y$.

@@

\begin{figure*}[t]
\[\begin{array}{rrcll}
\idc :& \tau & \iso & \tau &: \idc \\
\\
\identlp :&  0 + \tau & \iso & \tau &: \identrp \\
\swapp :&  \tau_1 + \tau_2 & \iso & \tau_2 + \tau_1 &: \swapp \\
\assoclp :&  \tau_1 + (\tau_2 + \tau_3) & \iso & (\tau_1 + \tau_2) + \tau_3 &: \assocrp \\
\\
\identlt :&  1 * \tau & \iso & \tau &: \identrt \\
\swapt :&  \tau_1 * \tau_2 & \iso & \tau_2 * \tau_1 &: \swapt \\
\assoclt :&  \tau_1 * (\tau_2 * \tau_3) & \iso & (\tau_1 * \tau_2) * \tau_3 &: \assocrt \\
\\
\distz :&~ 0 * \tau & \iso & 0 ~ &: \factorzl \\
\dist :&~ (\tau_1 + \tau_2) * \tau_3 & \iso & (\tau_1 * \tau_3) + (\tau_2 * \tau_3)~ &: \factor
\end{array}\]
\begin{center}
\Rule{}
{\jdg{}{}{c_1 : \tau_1 \iso \tau_2} \quad \vdash c_2 : \tau_2 \iso \tau_3}
{\jdg{}{}{c_1 \odot c_2 : \tau_1 \iso \tau_3}}
{}
\\
\Rule{}
{\jdg{}{}{c_1 : \tau_1 \iso \tau_2} \quad \vdash c_2 : \tau_3 \iso \tau_4}
{\jdg{}{}{c_1 \oplus c_2 : \tau_1 + \tau_3 \iso \tau_2 + \tau_4}}
{}
\qquad
\Rule{}
{\jdg{}{}{c_1 : \tau_1 \iso \tau_2} \quad \vdash c_2 : \tau_3 \iso \tau_4}
{\jdg{}{}{c_1 \otimes c_2 : \tau_1 * \tau_3 \iso \tau_2 * \tau_4}}
{}
\end{center}
\caption{$\Pi$-terms and combinators.}
\label{pi-terms}
\end{figure*}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{PiFrac: Generalizing Scoped Ancilla with Dynamic Checks}

@@

When ancilla bits are created they are created with a specific
value. This value is encoded in the circuit. The value of fractional
type represents the ``garbage collector.''  The garbage collector is
specialized to collect a single value, which is the value used to
create the ancilla value. At this point, if the garbage collector
encounters a different value, it doesn't know what to do, and would
need to throw an exception. This situation in \verb|PiFracDyn| is
similar to current solutions (e.g. Quipper) but actually more general
as it allows non-scoped ancilla allocation and de-allocation. Show
examples.

@@

New combinators:

\EtaEpsilon{}

and their evaluation:

\EtaEpsilonEval{}

%%%%%
\subsection{Examples}

\EtaEpsilonExamples{}

@@

Exploit dependent types to reify proofs in the type system. Type
checking requires (partial) eval. Once we have a proof can we somehow
go back to \verb|PiFracDyn| and remove the runtime check?
